Quantifying unmet ICU beds in tertiary hospitals in urban India using Propensity

### 2022-09-23 manuscript.Rmd
Update manuscript.Rmd

Abbreviations
============
ER - Emergency room
HIC - High-income countries
LMIC - Low- and middle-income countries

@@

<!-- 
FYI: utmanat volymkraven, de är motstridiga med kravet "Innehåller endast sådan information som är nödvändig för att förstå och motivera studiens syfte och metodik i stort"
-->

@@

Background
----------
Trauma/Injury <!-- Define trauma here, as there is a difference between trauma and Injury. Trauma can be defined as physical injury and the bodys physiological response --> is a substantial global health problem that disproportionately affects low and middle-income countries (LMIC), resulting in significant mortality and morbidity @traumaLMIC. Despite having only 1% of the global vehicle fleet, a disproportionate 10% of global traffic casualties occur in India @indiantrauma. The annual mortality from injury in India amount to one million casualties @onemilliondead. Half of trauma mortality in India occur in-hospital, and in-hospital trauma mortality in Indian hospitals was double that of high-income countries @indiantrauma.

Health care professionals in many LMIC testify that their medical assessment oftentimes is that a trauma patient should be admitted to the ICU, but aren’t because such resources are scarce, overloaded and unavailable. Their medical judgement is that the lives of these patients were salvageable, and that these patients died as a direct result of not being admitted to the ICU @stuffLMICdoctorssay. Globally, an estimated two million lives could be saved annually, if the appropriate and timely trauma care provided by hospitals in many high-income countries were available and provided in low to-middle-income countries (LMICs) @livessaved

In order to improve and ensure equivalence of intra-hospital care, trauma quality improvement programs (TQIP) have been developed. TQIP assessments collect trauma data from participating health care providers and apply statistical analysis to the data. The aim is to find discrepancies in treatment for patients presenting with similar conditions. TQIPs aim to ensure equivalent care <!-- I'm not sure equalivent care, more to improve the trauma care in general. --> and improve trauma care by developing best practice guidelines @tqip, and to assert that health care providers provide the same care under same circumstances as their peers do. TQIPs utilizes a retrospective cohort study, analysing the current of trauma care patients and post-fact attempting to determine the appropriate level of care for each patient. <!-- There are several types of interventions under the umbrella tern TQIP, ranging from, so not just one type of intervention. I'm not sure it can be refered to as a retrospective cohort study, as the processes for these interventions differ. --> From these analyses many aspects of trauma care can be improved and optimized, such as determining the appropriate amount of ICU beds and other trauma care resources @tqip. <!-- As far as I know estimating number of ICU beds has not been explicity done within TQIP, likely it is done base on caseload in general health planning and management in HIC, I would suggest removing the last part of that sentence -->
Fråga: rätt uppfattat? bygger dessa analyser alltid på propensity score matching approach ? <!-- No, I would say that is rare... -->

<!-- I suggest adding more details about the context in India, the health care system, how common trauma is in India, how many ICU beds there are, why ICU care is important for the trauma population and so on.. -->

Knowlege gap
------------
Existing TQIP often originate in high-income countries (HIC), and are either not applicable nor easily transferred to the hospital conditions of low and middle income countries <!-- you can use LMIC here, as you've previously defined it --> (LMIC) @frailty @gunshot  @onemilliondead. Applying TQIPs outside of their original context or even within a subset of it's own origin is difficult even without taking income bracket in account, for instance firearm violence is a considerable bigger portion of trauma patients in some US contexts, than the national average @gunshot. For trauma patients, there exist significant structural differences between HIC and LMIC. These differences are in factors such as injury mechanisms and severity, health care resources as well as patient make-up @frailty @indiantrauma @traumaLMIC @onemilliondead. One example of this is the differences in helmet use for motorcyclists and their passengers, as well as differences of displacement in vehicle fleets in LMIC versus HIC. These factors have tremendous impact on the type and severity of injuries patients have, and subsequently on their care and outcome. These differences make a direct transferral of HIC TQIP models onto LMIC irrelevant or misleading and there is a need to develop TQIP assessment applicable to local conditions. <!-- This is true, however, your thesis is on specificlly estimating the need for ICU beds for this trauma population in India, and not technincally a TQIP so I would suggest focusing less on this and more on the trauma population, the need for ICU care and why we belive there is a lack of ICU beds in this setting. This is the knowledge gap, especially a quantifiable number of this shortage of beds. -->

Trauma is a main cause of mortality and morbidity to otherwise healthy and young individuals, and dispropotionately affect this demographic. People in this demographic are also oftentimes the primary or sole breadwinners for their families. Therefore the detrimental impact of trauma go beyond mortality or substantial loss of disease-free years of life for the affected individuals, it also carries a considerable financial and social harm on their family members and communities @traumaLMIC. <!-- This is true and an important aspect to bring up in the background, this section however does not explain the knowledge gap. -->

Aim
----
Predict the unmet need for ICU beds for trauma patients at tertiary hospitals in urban India.

This study aims to make an TQIP <!-- I would avoid calling it a TQIP--> assessment in an Indian context to fill part of the knowledge gap, and to test whether there is merit in the testimony of the Indian colleagues. The TQIP assessment shall propose a scope of the unmet need for ICU beds for trauma patients at tertiary hospitals in urban India. If this analysis indeed indicate a shortage, it can be used to motivate extending ICU resources and reduce preventable trauma morbidity and mortality.

<!-- This is more about methods.. --> The theory being examined in this masters thesis is that there exist subset of the TITCO dataset that didn't go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to receive this intervention. The thesis is that these patients were not admitted to the ICU only for lack of such resources, not by a medical assessment. In statistics, such a group would be akin to an non-experimental conrol group, and these patients will be identified using propensity score matching.

The size of this group in relation to the group that were admitted to the ICU indicate the magnitude of the unmet need of ICU beds. If, for every patient that did receive ICU care, there was another patient with similar predictors that wasn’t admitted due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be. <!-- I think these three sections can be moved or removed from this section, the aim is enough -->

@@

<!-- Present key elements of study design -->
Observational retrospective cohort study from data collected at four tertiary hospitals in urban India.

@@

<!-- Describe the setting, locations, and relevant dates, including
periods of recruitment, exposure, follow-up, and data collection -->
16 000 patients admitted to four public university hospitals in urban India are included in the TITCO<!-- add reference to the dataset as well--> dataset. Initially the data set included 16047 patients, but to improve anonymization 47 patients were deleted from the dataset as described in the chapter anonymization. The participating hospitals were King Edward Memorial hospital (KEM) and Lokmanya Tilak Municipal General Hospital (LTMGH), Mumbai, Jai Prakash Narayan Apex Trauma Center (JPNATC), All India Institute of Medical Sciences, New Delhi and The Institute of Post-Graduate Medical Education and Research and Seth Sukhlal Karnani Memorial Hospital (SSKM), Kolkata. When the data was collected all hospitals were tertiary hospitals. JPNATC had almost 180 beds in a dedicated trauma centre, KEM had no specific trauma ward, LTMGH was a public university hospital with a specific trauma ward with 14 beds and SSKM was a  public university hospital with no specific trauma ward.

@@

<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->
The eligibility criteria for TITCO were patients who presented with trauma at the emergency departments of the participating hospitals. Trauma covers injury and with a mechanism of road traffic, railway, fall, burn or assault. Patients who were dead on arrival were excluded, as were patients who only had isolated limb injuries.
A majority of patients came from low socioeconomic circumstances in their respective communities.

@@

<!-- Cross-sectional study: Give the eligibility criteria, and the
sources and methods of selection of participants -->
Ethical considerations
----------------------
<!-- This is the ethical approvals for collecting the TITCO data right? I suggest writing that clearly and that our study does not requlre ethical permission. I'm not sure this is needed, to have the ethical permissions for the TITCO study as we are refering to it and using the data collected and shared openly, but we can keep it for now! --> Ethical approval for the collection of the data was obtained from the boards of ethical approval in all four hospitals.

@@

FRÅGA: Kan man och är det meningsfullt att analysera om sjukhus med dedikerade trauma wards får fler patienter körda till dem som sen hamnar på ICU? För när jag tittade lite på om man kunde luska ut vilket sjukhus som var vilket via just tran-parametern så var skillnaden tydlig. På ett sjukhus hamnar 96% av patienterna som transporterats från andra health facilities på IVA, och på ett annat sjukhus är det bara 4% och då gissar jag att det förra är ett sjukhus med en dedikerad trauma ward och det senare inte är det?
det är vad det är för anonymiseringen det är lite begränsade möjligheter att anonymisera bra som jag illustrerar med järnvägsexemplet, men jag ställer frågan för om detta är en bias jag borde ta hänsyn till eller iallafall acknowledge ? Jag ställer i princip samma fråga i bias-kapitlet. <!-- Hm, jag tänker att vi inte ska ha med variabeln centre i modellen, för den kommer väga för tungt/korrelera för bra med outcome i det fallet med ett sjukhus som har så hög inläggningsgrad på IVA för den patienten. Exakt varför det ser ut så vet jag inte, det kan säkert Martin svara på som kan kontexten där datat hämtats in bättre. -->

@@

Date observations
Dates recorded in the TITCO data set were shifted using a random offset of somewhere between 0 and 5 weeks for each patient, but without losing any relevant information. The offset was designed to not dissipate details that could be of potential use for researchers such as day of week and rough indication of season. The internal relations between dates have been preserved so that the timing of all recordings and discharge in chronological relation to admission are all intact.

It is important to note that all patients have sequence numbers that corresponded to the chronological order in which they arrived at any of the participating hospitals, and that sequence number has been kept intact. Keeping this sequence number variable intact allows for many useful operations such as examinations of trends in the data, while it is not possible to undo the date randomization by the sequence number variable, such an equation system is still underdetermined. Unfortunately this means it is also not possible to reconstruct the patient volumes at any given hospital on any given day and from that rate of resource engagement. Nor is it possible to distinguish if patients who were made to wait for their first assessment were so because their injuries were deemed as less severe or if it was because health care staff was overwhelmed which is an important distinction. <!-- Good that you've included this! -->

@@

Age
Because very young and very old patients are so few, they have been grouped together to prevent identification via the age parameter.

Fråga: man har buntat ihop 89+ arna för anonymisering, men 0-åringarna är bara 7 och ganska sårbara indentifieringsmässigt <!-- Jag skulle tro det ja! -->

@@

<!-- Describe any efforts to address potential sources of bias -->
The TITCO-I dataset are observational data collected from a clinical environment and not an experimental setting.  Medical interventions, examinations and ICU admission are not randomized, they are a consequence of a patients medical condition. There exists a causal effect between the independent and the dependent variables: a patient is sent to the ICU precisely because of the severity of their condition.

Because the purpose of this thesis is to statistically assess how many patients that should have gone to the ICU but didn't, this particular selection bias does not skew the outcome and therefore does not have to be adressed.

FRÅGA: Bör jag tänka på någon bias iallafall? Jag kan tänka mig att det finns skillnader mellan sjukhusen som jag skriver lite om innan (de har väl inte nödvändigtvis exakt samma proportioner ICU-sängar givet upptagningsområde, och att ett sjukhus därför kan vara mycket mer generös med ICU-sängarna än ett annat? ICU admission rate skiljer sig ju en del och sjukhus med dedikerade trauma wards får fler svårt skakade patienter - inte minst ditkörda från andra sjukhus - som åker in på ICU, än sjukhus utan dessa resurser. en trauma patient med viss karakteristik påkörd utanför sjukhus A vs utanför sjukhus D verkar göra skillnad.), men vad för typ av statistisk bias är det isåfall? 
Det finns en selection bias i att det är lägre socioekonomisk stratum, men hur relevant är den? Och om detta gäller hela data setet så påverkar det väl mera hur generaliserbara resultaten är, snarare än introducerar en bias jag måste addressera? <!-- Håller med, jag tänker att de olika sjukhusen absolut inducerar en bias. När det gäller socioekonomi håller jag med, det handlar mer om att de här sjukhusen är publika sjukhus i indien och har mer med generaliserbarhet att göra. -->

Propensity score matching in itself introduces a source of bias. The researcher chooses the predictor variables and so can introduce an overdetermined set and a self-fulfilling prophecy. For this reason the dataset shall be randomly split into two parts of 40% and 60% of the patients. The 40% shall be used to produce the propensity score matching, and the 60% shall be used to verify it. <!-- I'm not sure it's a bias, more that how the method is applied can effect the results and how many patient that match in the end.. We will train on 60% of the dataset and validate/test in the 40%! -->
Fråga: typ?

@@

<!-- Explain how the study size was arrived at -->
The data of 16047 trauma patients were collected at the participating hospitals over 2.5 years, as described in chapter Settings.
47 entries have been randomly excluded for anonymization purposes.

Återkom: XX patients were discarded because of missing data <!-- This is for the results! -->

Quantitative variables
----------------------

<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->
Fråga: hur diffar detta från vad som finns i Variables and data sources/measurements? Har jag missförstått rubriken variables eller den här rubriken? <!-- Since you described this elsewhere you can remove it I'd say -->

Statistical methods
-------------------

(a) Describe all statistical methods, including those used to control
for confounding
-->
Propensity score matching will be used in this analysis. Propensity score matching (PSM) is best explained using one typical application: identifying a control group when examining the efficacy of an intervention for a specific outcome, where there is causality between the independent variables for the intervention, and sometimes also the outcome. There is a need to offset or balance the effect of this causality, but a randomized case-control study could be impractical or immoral.

In this context, it could be to examine the efficacy of ICU care on survival. If the correlation between the independent variables and the intervention and outcome isn't accounted for, a simple comparison between the cohort that did go to the ICU to the cohort that did not, would find that ICU care actually increased mortality. In the TITCO-I dataset the survival rate of patients who did not go to the ICU was 81% versus 68% for patients who did. This is of course an incorrect conclusion, this data has treatment selection bias as patient covariates often confound treatment selection. The higher mortality observed in patients who received ICU care is because patients were selected for the intervention because they presented certain medical conditions that also carry a high mortality - even with the intervention ICU care. Producing a randomized control trial study for these types of lifesaving medical interventions for trauma patients would obviously go against ethical guidelines of research, and another approach to produce a control group is necessary. Propensity score matching is a statistical procedure for reducing this bias by assembling a sample from within the data set, in which confounding factors are balanced between the cohort that got the intervention and the cohort that did not. <!-- I think this example on how PSM is typically used may not be nessicary, it is more important to describe the steps we take in the method. -->
<!-- 
The goal of propensity score matching is usually to identify suitable control subjects that did not receive an intervention, even though they fairly well met the same inclusion criteria that the subjects that did receive the intervention had. Control subjects who are similar enough to the intervention group in the most relevant aspects for receiving treatment. 
-->
The size of this subset in relation to the subset of patients that did receive ICU care is the end point for this analysis. This proportion is indicative of the unmet need of ICU beds, as described in chapter Aim.
In the typical PSM appliction, this subset is used as the control group to determine the efficacy of ICU care for some outcome, eg mortality.

<!-- We first split the data, 60-40% -->
The first step is to identify predictor variables: covariates most predictive of the dependent variable. In this context it means to first identify the conditions and medical interventions received that correlate best to being offered ICU care. All properties of the patients that were sent to the ICU will be compared to the patents who didn't, in order to identify properties that distinguishes this cohort. <!-- We will use both the data as a source for this as well as medical expertise to select covariates. We will also avoid covariates with a correlation close to 1 to avoid overfitting. -->

Second step is to calculate propensity scores <!-- This is the same as the probability from the logistica regression--> for all patients. Each patient in the TITCO data set gets a propensity score based on their medical conditions and interventions, specifically their individual values of the predictor variables. The propensity score determines the likelihood that a patient will receive ICU care, given their condition.

Third step is PSM. PSM is a standardized mathematical approach to identify members of the subset that did not get the intervention but had the same predictors as the subgroup that did get the intervention. The purpose of the propensity score matching is identifying patients that weren’t admitted into the ICU, but presented with a set of predictors - a propensity score - very similar to those that were admitted.

@@

The two subgroups were patients admitted to the ICU, and the patients who weren't. Admission to the iCU is a binary yes or no, length of stay has not been taken into Consideration.

Missing Data
-------------
The data completion rate is presented in table_nums('tab_1'). As the base assumption, all quantitative variables collected for each patient are considered potential recordings. Data that are missing are assumed missing because care for a patient did not motivate the collection of the data. For example, if serum creatinine was not recorded, it is assumed that the collection of this variable wasn't relevant for care of the patient.

@@

All quantitative data from the original dataset that were entered as zero but where a zero value is physiologically impossible were replaced with NA values. Zero values for variables where it is possible to have zero values, such as respiration rate and heart rate were left as zeroes to not tamper with the data, although a zero value indicate that the variable probably wasn't measured. These amount to 250 and 155 entries respectively for the first observation within an hour and less than 100 for both for the second observation after 24 h, so the choice to include or exclude these observations won’t skew the data considerably.

@@

The TITCO-I data set is analyzed mathemathically in pursuit of suitable predictor variables. The data set is divided into two subgroups, the cohort that were admitted to the ICU and the cohort that wasn't. In comparison between these subgroups, identify variables that distinguishes the former from the latter. The quantitative and qualitative properties were considered separately.

@@

The quantitative data results see table_nums('tab_2')

>>>>

Abbreviations
============
ACS - American college of surgeons
ER - Emergency room
HIC - High-income countries
LMIC - Low- and middle-income countries
PSM - propensity score matching
RCT - Randomized case-control study
TITCO -Towards Improved Trauma Care Outcomes 
TQIP - Trauma Care Improvement Program

@@

@@

Trauma globally and in India
----------------------------
Trauma is a substantial global health problem that disproportionately affects low and middle-income countries (LMIC), resulting in significant mortality and morbidity @traumaLMIC. Despite having only 1% of the global vehicle fleet, a disproportionate 10% of global traffic casualties occur in India @indiantrauma. The annual mortality from injury in India amount to one million casualties @onemilliondead. Half of trauma mortality in India occur in-hospital, and in-hospital trauma mortality in Indian hospitals was double that of high-income countries @indiantrauma.

Mortality due to accidents in India amounted to 421 104 people in 2019, which is 31.5 per 100.000 inhabitants. The traumatic mechanisms of injury from accidents covered in this thesis are road accidents, falls, assault and burns and these make up about half of the accident mortality @nationalcrime.

Trauma is a main cause of mortality and morbidity to otherwise healthy and young individuals, and dispropotionately affect this demographic. Most casualties affected the age group 30-45 years, followed by the age bracket 18-30 years @nationalcrime. People in this demographic are also oftentimes the primary or sole breadwinners for their families. Therefore the detrimental impact of trauma go beyond mortality or substantial loss of disease-free years of life for the affected individuals, it also carries a considerable financial and social harm on their family members and communities @traumaLMIC @nationalcrime

Health care professionals in many LMIC testify that their medical assessment oftentimes is that a trauma patient should be admitted to the ICU, but are not because such resources are scarce, overloaded and unavailable. Their medical judgement is that the lives of these patients were salvageable, and that these patients died as a direct result of not being admitted to the ICU @stuffLMICdoctorssay. Globally, an estimated two million lives could be saved annually, if the appropriate and timely trauma care provided by hospitals in many high-income countries were available and provided in low to-middle-income countries (LMICs) @livessaved

A diverse mix of public, for-profit and not-for-profit private actors provide health services in India. According to a WHO 2022 report, India has 76 709 non-ventilator supported and 39 476 ventilator-supported ICU beds @ICUbeds. This is 8.2 per 100 000 inhabitants.
The WHO report also highlight the probable shortage of ICU resources in India, though in the context of the COVID-19 pandemic.

Methods to examine trauma care and ensure equivalence
-----------------------------------------------------
In order to improve and ensure equivalence of intra-hospital care, trauma quality improvement programs (TQIP) have been developed. The pilot and most notable example is the TQIP developed by the American College of Surgeons ACS. Since the implementation of the ACS pilot, TQIP has evolved to a broad term that include many kinds of processes to improve the quality of trauma care, such as mortality and morbidity conferences, using audit filters or comparing risk adjusted outcomes over time and between centres. The aim is to find discrepancies in treatment for patients presenting with similar conditions. TQIPs aim to ensure patients get equivalent care within and across hospitals, to benchmark performance of health care providers versus their peers and to improve trauma care by developing best practice guidelines @tqip.

One aspect of TQIP assessments is to collect trauma data from participating health care providers and apply statistical analysis to the data. From this data, TQIPs utilizes a retrospective cohort study, analysing the current of trauma care patients and post-fact attempting to determine the appropriate level of care for each patient. From these analyses many aspects of trauma care can be improved and optimized, such as determining the appropriate amount of ICU beds and other trauma care resources @tqip.

It is tempting to apply existing models of TQIP assessments on Indian trauma data to identify causes of the observed disparities in trauma care performance in HIC and LMIC, and propose measures to harmonize performance. Unfortunately as TQIP often originate in high-income countries (HIC), they are either not applicable nor easily transferred to the hospital conditions of LMIC @frailty @gunshot  @onemilliondead. Applying TQIPs outside of their original context or even within a subset of its own origin is difficult even without taking income bracket in account, for instance firearm violence is a considerable bigger portion of trauma patients in some US contexts, than the national average @gunshot. For trauma patients, there exist significant structural differences between HIC and LMIC. These differences are in factors such as injury mechanisms and severity, health care resources as well as patient make-up @frailty @indiantrauma @traumaLMIC @onemilliondead. One example of this is the differences in helmet use for motorcyclists and their passengers, as well as differences of displacement in vehicle fleets in LMIC versus HIC. These factors have tremendous impact on the type and severity of injuries patients in the ER have, and subsequently on their care and outcome. These differences make a direct transferral of HIC TQIP models onto LMIC irrelevant or misleading and there is a need to develop TQIP assessment applicable to local conditions.

Kommentar på Mgerdin kommentar: jag snodde din definition här. Jag håller med om att detta inte är den mest relevanta informationen men för "kontrollmekanism"-arbetet som ska in på måndag här har KI en checklista och de vill ha 3-5 sidor bakgrund trots att vi jobbar med max 1-1.5 så jag måste feta ut. Detsamma för kommentaren i Aim jag är medveten om att det är felplacerat eller irrelevant.

Knowlege gap
------------
There are compelling reasons to believe there is a shortage of ICU beds in India, but the scope of this shortage and the potential for reduction of trauma mortality if this shortage would be remedied is unknown.

WHO and local health care workers believe there is a shortage of ICU beds in Indian hospitals, and that the lives of many patients would have been salvageable if the appropriate ICU resources had been available. The differences of in-hospital trauma mortality in Indian hospitals was double that of high-income countriesin, and part of this difference could be explained by a shortage of ICU beds in India @indiantrauma

Aim
----

Quantify the unmet need for ICU beds in tertiaty hospitals in urban India.

Apply statistical analysis to a large data set of Indian trauma patients and determine how many out of the patients that were not admitted to the ICU reasonably should have been admitted given their condition. Their numbers in proportion to the number of ICU-admitted patients is indicative of the relative shortage of ICU beds.

@@

<!-- Present key elements of study design -->
Observational retrospective cohort study using data from the Towards Improved Trauma Care Outcomes (TITCO) in India cohort.

@@

<!-- Describe the setting, locations, and relevant dates, including
periods of recruitment, exposure, follow-up, and data collection -->
16 000 patients admitted to four public university hospitals in urban India are included in the TITCO dataset @titco. Initially the data set included 16047 patients, but to improve anonymization 47 patients were deleted from the dataset as described in the chapter anonymization. The participating hospitals were King Edward Memorial hospital (KEM) and Lokmanya Tilak Municipal General Hospital (LTMGH), Mumbai, Jai Prakash Narayan Apex Trauma Center (JPNATC), All India Institute of Medical Sciences, New Delhi and The Institute of Post-Graduate Medical Education and Research and Seth Sukhlal Karnani Memorial Hospital (SSKM), Kolkata. When the data was collected all hospitals were tertiary hospitals. JPNATC had almost 180 beds in a dedicated trauma centre, KEM had no specific trauma ward, LTMGH was a public university hospital with a specific trauma ward with 14 beds and SSKM was a  public university hospital with no specific trauma ward.

The participating hospitals are all public entities, so a majority of patients came from low socioeconomic circumstances in their respective communities.

@@

<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->
The eligibility criteria for TITCO were patients who presented with injuries from trauma at the emergency departments of the participating hospitals with a mechanism of road traffic, railway, fall, burn or assault. Patients who were dead on arrival were excluded, as were patients who only had isolated limb injuries.

@@

<!-- Cross-sectional study: Give the eligibility criteria, and the
sources and methods of selection of participants -->
Ethical considerations
----------------------

This study does not requlre ethical permission, it refers to data collected and shared openly.

Ethical approval for the collection of the data used was obtained from the boards of ethical approval in all four hospitals.

@@

@@

Date observations
Dates recorded in the TITCO-I data set were shifted using a random offset of somewhere between 0 and 5 weeks for each patient, but without losing any relevant information. The offset was designed to not dissipate details that could be of potential use for researchers such as day of week and rough indication of season. The internal relations between dates have been preserved so that the timing of all recordings and discharge in chronological relation to admission are all intact.

It is important to note that all patients have sequence numbers that corresponded to the chronological order in which they arrived at any of the participating hospitals, and that sequence number has been kept intact. Keeping this sequence number variable intact allows for many useful operations such as examinations of trends in the data, while it is not possible to undo the date randomization by the sequence number variable, such an equation system is still underdetermined. Unfortunately this means it is also not possible to reconstruct the patient volumes at any given hospital on any given day and from that rate of resource engagement. Nor is it possible to distinguish if patients who were made to wait for their first assessment were so because their injuries were deemed as less severe or if it was because health care staff was overwhelmed which is an important distinction.

@@

Age
Very old patients are few, they have been grouped together to prevent identification via the age parameter.

@@

<!-- Describe any efforts to address potential sources of bias -->
The TITCO dataset are observational data collected from a clinical environment and not an experimental setting.  Medical interventions, examinations and ICU admission are not randomized, they are a consequence of a patients medical condition. There exists a causal effect between the independent and the dependent variables: a patient is sent to the ICU precisely because of the severity of their condition.

There exist a potential bias in the routines, resources and patient makeup of the participating hospitals. Becasue they are all tertiary hospitals, about two thirds of the patients have been referred from other health care providers, meaning that there is a selection bias in the patient populations. Further, hospitals with dedicated trauma wards have a propensity to receive transferred patients or police escorted patients with worse condition and prognosis.

As stated in chapter Settings, most patients have low socioeconomic status. Low socioeconomic status is associated with higher level of malnutrition and comorbidities and would have a negtive effect on vitals and the capacity of patients to tolerate certain interventions. This circumstance doesn't affect the analysis and internal validity, but reduce the generalizability of the results. 
<!-- (ÅTERKOM: har inte råd att refereas?) -->

Because the purpose of this thesis is to statistically assess how many patients that should have gone to the ICU but did not, these particular selection biases are not expected to skew the outcome, and therefore does not have to be adressed.

Hospitals with trauma wards could have lower admission criteria to the ICU than hospitals with no trauma ward. Because examining admission criteria to the ICU is part of the analysis it may be necessary to examine and address this bias by comparing ICU admitted patients between hospitals. Propensity score matching can be used to identify differences in ICU admission criteria between hospitals and find patients that were admitted but need not have been. These patients are excluded from predictor selection to reduce this bias. This is part of the development and training of the PSM model.

Propensity score matching in itself introduces a source of bias from unfit assumptions. The terminology and potential sources of bias are explained chapter Study Design. In choosing predictor variables the researcher could introduce an overdetermined set and a self-fulfilling prophecy. For this reason the dataset shall be randomly split into two parts of 40% and 60% of the patients. The 60% shall be used to produce the propensity score matching, and the remaining 40% shall be used to verify it.

@@

<!-- Explain how the study size was arrived at -->
The data of 16000 trauma patients were collected at the participating hospitals over 2.5 years, as described in chapter Settings.

Statistical methods
-------------------
<!--

(a) Describe all statistical methods, including those used to control
for confounding
-->
Propensity score matching (PSM) is useful method for identifying and assembling a control group from within a data set where there is causality between the independent and dependent variables, and a randomized case-control (RCT) study to offset or balance this bias is impractical, immoral or otherwise improper. Propensity score is a statistical variable, understood as the likelihood of a subject receiving treatment conditional on observations of relevant covariates of that subject before treatment. In RCTs it is known from study design, but it is generally unknown in observational data. The rationale for PSM is that while it is unknown, this variable can be fairly well estimated using logistic regression.

The theory being examined in this masters thesis is that there exist a subset of the TITCO dataset that did not go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to receive this intervention. The thesis is that these patients were not admitted to the ICU only for lack of such resources, not by a medical assessment.

Translated into this context this means that in this data set exist a selection bias, there is causality between the independent variables medical conditions and interventions and the dependent variable ICU admittance. Patients were admitted to the ICU as a consequence of their medical condition. Performing a randomized case-control study and randomly assigning or withholding a live-saving treatment would obviously be immoral, but we believe that in the TITCO data set exist patients who were not admitted into the ICU by volition but for scarcity of such resources. Such a group would be akin to an non-experimental control group and PSM can be used to identify and assemble it.  If such a subset can be identified by PSM, then it naturally follows that there exists shortage of ICU beds in tertiary hospitals in India. It also follows that the shortage can be quantified by the proportionality of this subset versus the subset that did receive ICU care. If, for every patient that did receive ICU care, there was another patient with similar predictors that wasn’t admitted due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.

The propensity score model is developed using the Training Data and tested using the Validation Data. The workflow contain a series of steps and the work flow should understood as linear, but iterative. The output of each step is the input for later steps. Therefore the output of the last step reflect the cumulative error from unfit assumptions made the second and fourth step. The output is tested against Validation Data. By reverting back and adjusting assumptions in an iterative process, the model is continuously improved and the accuracy and validity is improved.

The first step of PSM is to randomly split the data set into two parts, 60% and 40%. The 60% part is a Training Part used to delvelop and train the PSM, and the 40% part is the Validation Part. This is done because the inherent bias of the PSM method described in chapter Bias. Training Data is used in steps 2-4 and the output is then validated using Validation Data.

Second step is to identify predictor variables: covariates most predictive of the dependent variable, as described in chapter Predictor selection. In this context it means to first identify the conditions and medical interventions received that correlate best to being admitted to ICU care. This selection is based on clinical experience as well as analysis of the Training Data. In the data analysis all properties of the patients that were sent to the ICU will be compared to the patents who did not, in order to identify properties that distinguishes this cohort., Familiarity with the collection of source data is crucial to minimize the risk to choose irrelevant predictors and introduce error.

Third step is to build a standard multivariable logistic regression model with ICU care as the outcome and the predictors selected in the second step as independent variables. This model is used to calculate the standard variable propensity score for all patients. The propensity score determines the likelihood that a patient will receive ICU care, given their condition.

Fourth step is PSM. In PSM members of the subset that did not get ICU care but had similar predictors - and therefore similar propensity scores - as the subgroup that did get ICU care, are matched to patients who did go to the ICU. There are a few different ways to perform this matching but they are all imperfect because propensity scores are unique and perfect matching is unattainable There are control mechanisms and limitations but a risk of applying too fix or too loose matching criteria the outcome could inadvertedly vastly under- or overestimate the  need of ICU beds remain.

The same procedure is repeated on the Validation Data using the same predictors and the same matching criteria, and this should produce a cohort comparable to the one trom Training Data.

@@

The two subgroups were patients admitted to the ICU, and the patients who were not. Admission to the iCU is a binary yes or no, length of stay has not been taken into Consideration.

Missing Data
-------------
The data completion rate is presented in table_nums('tab_1'). As the base assumption, all quantitative variables collected for each patient are considered potential recordings. Data that are missing are assumed missing because care for a patient did not motivate the collection of the data. For example, if serum creatinine was not recorded, it is assumed that the collection of this variable was not relevant for care of the patient.

@@

All quantitative data from the original dataset that were entered as zero but where a zero value is physiologically impossible were replaced with NA values. Zero values for variables where it is possible to have zero values, such as respiration rate and heart rate were left as zeroes to not tamper with the data, although a zero value indicate that the variable probably was not measured. These amount to 250 and 155 entries respectively for the first observation within an hour and less than 100 for both for the second observation after 24 h, so the choice to include or exclude these observations won’t skew the data considerably.

@@

The TITCO-I data set is analyzed mathemathically in pursuit of suitable predictor variables. The data set is divided into two subgroups, the cohort that were admitted to the ICU and the cohort that was not. In comparison between these subgroups, identify variables that distinguishes the former from the latter. The quantitative and qualitative properties were considered separately.

@@

Missing data
------------
Återkom: XX patients were discarded because of missing data

Predictor selection
-------------------
The quantitative data results see table_nums('tab_2')
